\large How resource abundance and stochasticity affect organisms' range sizes

### 2024-05-30 writing/appendix-C-empirical-modeling.Rmd
updated writing

---
title: \large How resource abundance and stochasticity affect animals' space-use requirements
subtitle: "Appendix C: Empirical modeling"
# author:
#   - name: Stefano Mezzini
#     email: stefano.mezzini@ubc.ca
#     institute: [biol, braes]
#   - name: Chris H. Fleming
#     institute: [ucf, smith]
#   - name: E. Patrícia Medici
#     institute: [tapir, iucn, escas]
#   - name: Michael J. Noonan
#     email: michael.noonan@ubc.ca
#     institute: [biol, braes, cmps]
# institute:
#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
header-includes:
    - \renewcommand{\figurename}{Figure C\!\!} # for "Figure Sx.
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize} # smaller caption text
    -  \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
---

@@

We start by providing examples and considerations on how one may model $R$ and the effects of $\e(R)$ and $\var(R)$ on organisms' space-use requirements. Next, we apply the methods to the tapir's GPS tracking data and use NDVI as a proxy for $R$. Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating organisms' space-use requirements under different scenarios.

@@

\noindent such that an organism's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec l_t$), as well as how each of the parameters may change in response to the current environmental factors ($\vec \xi_t$). This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible). The parameters $\Omega, \Phi, \vec w_t$ and $\vec \xi_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., limb length, body size, site fidelity, cortisol levels, reproductive stage), doing so with sufficient frequency and accuracy would not be simple. Thus, we can instead focus on organisms' movement and space-use requirements as a response to overall resource abundance and unpredictability while accepting other drivers of movement as causes of stochastic behaviors.

@@

Accurate speed estimates require the sampling interval, $\Delta t$, to be sufficiently small [$\Delta t < \tau_v$, see @noonan_scale-insensitive_2019], since the data must contain multiple consecutive samples where the organism is moving with a similar speed and direction. However, since the time required for an organism to cross its home range is orders of magnitude larger than the time it moves with a consistent behavior (i.e., $\tau_p > \tau_v$), it is still possible to produce good estimates of range crossing time even if there is no information on the organism's speed. Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the organism's space-use requirements and home-range size, provided that the total sampling duration, $T$, is sufficiently long. Model selection procedures with criteria such as the Akaike Information Criterion [@akaike_new_1974] can help choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].

@@

### Quantifying space-use requirements

\noindent The concept of a home range has a long history in ecological research as an indicator of the space an organism requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for organisms which are not range-resident, such as migrating organisms [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic organisms [e.g., @morato_space_2016; @nandintsetseg_variability_2019]. Although here we quantify organism's space-use requirements using 95% home-range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of organisms that are not range-resident. In the case of organisms that are range resident for short periods of time, we can model their movement and short-term space-use requirements within distinct and discrete periods or using a sliding window approach (see the discussion section). Note, however, that although range-resident organisms will generally have smaller positional variance ($\vpos$) than migratory and nomadic organisms, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an organism's movement but not its behavior. Instead, the organism's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different behaviors. Organisms with a $\tau_p$ shorter than a year are likely range-resident, while organisms with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year). In contrast, when an organism's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it does not have an average position it returns to regularly.
-->

<!-- ### Modeling space-use requirements -->

@@

This section of the appendix illustrates the steps necessary to reproduce the tapir movement analysis and the related figure in the manuscript (fig. 5). The tapir data used here is from the work of @medici_movement_2022 and can be found at the GitHub repository located at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs). Estimating the effects of resource abundance and unpredictability on the tapir's space-use requirements requires us to first estimate the changes in the tapir's space-use requirements (section \@ref(movement)) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi)) before we can estimate the relationship between resource dynamics and space use (section \@ref(final-fig)).

@@

```{r, eval=TRUE, warning=FALSE}
# attach all necessary packages
library('terra')     # to import and save rasters
library('dplyr')     # for data wrangling
library('purrr')     # for functional programming
library('tidyr')     # for data wrangling
library('ggplot2')   # for fancy plots
library('cowplot')   # for fancy multi-panel plots
library('ctmm')      # for movement modeling
library('mgcv')      # for empirical Bayesian GAMs
library('lubridate') # for smoother date wrangling
library('sf')        # for spatial features
library('MODIStsp')  # for downloading NDVI rasters
library('dagitty')   # for directed acyclical graphs
library('ggdag')     # for directed acyclical graphs
library('gratia')    # for ggplot-based GAM figures
theme_set(theme_bw()) # change default theme
source('../functions/betals.r') # betals family written by Simon Wood
source('../analysis/figures/default-figure-styling.R') # for color palettes
source('../earthdata-login-info.R') # personal login info for EarthData
source('../functions/window_hr.R') # function to calculate HRs
```

@@

# axis labels (unicode characters removed because they knitting errors)
e_r <- bquote(paste(bold('Resource abundance')))
v_r <- bquote(paste(bold('Resource stochasticity')))
hr_lab <- bquote(bold(paste('7-day space-use requirement (', km^2, ')')))

@@

p_f <-
  expand_grid(mu = seq(from = floor(min(tapir$mu) * 100) / 100,
                       to = ceiling(max(tapir$mu) * 100) / 100,
                       length.out = 250),
              sigma2 = seq(from = 0, to = 0.0035, length.out = 250)) %>%
  mutate(hr_full_est = predict(m, newdata = ., type = 'response')) %>%
  ggplot() +
  geom_raster(aes(mu, sigma2, fill = hr_full_est)) +
  geom_contour(aes(mu, sigma2, z = hr_full_est), color = 'black') +
  geom_point(aes(mu, sigma2), tapir, alpha = 0.3, show.legend = FALSE) +
  scale_x_continuous(e_r, expand = c(0, 0)) +
  scale_y_continuous(v_r, expand = c(0, 0)) +
  scale_fill_gradient(bquote(atop(bold('7-day space-use'),
                                  paste(bold('requirement (km'), '\U00B2',
                                        bold(')')))),
                      low = 'grey90', high = pal[3], limits = c(0, NA)) +
  theme(legend.position = c(1, 1),
        legend.justification = c('right', 'top'),
        legend.box.background = element_rect(),
        legend.background = element_rect(),
        legend.key.width = unit(0.35, 'in')) +
  guides(fill = guide_colorbar(
    title.position = 'top',
    theme = theme(legend.title = element_text(hjust = 1)),
    direction = 'horizontal'))

>>>>

---
title: \large How resource abundance and stochasticity affect organisms' range sizes
subtitle: "Appendix C: Empirical modeling"
# author:
#   - name: Stefano Mezzini
#     email: stefano.mezzini@ubc.ca
#     institute: [biol, braes]
#   - name: Chris H. Fleming
#     institute: [ucf, smith]
#   - name: E. Patrícia Medici
#     institute: [tapir, iucn, escas]
#   - name: Michael J. Noonan
#     email: michael.noonan@ubc.ca
#     institute: [biol, braes, cmps]
# institute:
#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
# csl: 'the-american-naturalist.csl'
fontsize: 12pt
header-includes:
    - \renewcommand{\figurename}{Figure C\!\!} # for "Figure Sx.
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize} # smaller caption text
    -  \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
---

@@

We start by providing examples and considerations on how one may model $R$ and the effects of $\e(R)$ and $\var(R)$ on organisms' home-range sizes. Next, we apply the methods to the tapir's GPS tracking data and use NDVI as a proxy for $R$. Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating organisms' range sizes under different scenarios.

@@

\noindent such that an organism's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec l_t$), as well as how each of the parameters may change in response to the current environmental factors ($\vec \xi_t$). This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible). The parameters $\Omega, \Phi, \vec w_t$ and $\vec \xi_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., limb length, body size, site fidelity, cortisol levels, reproductive stage), doing so with sufficient frequency and accuracy would not be simple. Thus, we can instead focus on organisms' movement and range sizes as a response to overall resource abundance and unpredictability while accepting other drivers of movement as causes of stochastic behaviors.

@@

Accurate speed estimates require the sampling interval, $\Delta t$, to be sufficiently small [$\Delta t < \tau_v$, see @noonan_scale-insensitive_2019], since the data must contain multiple consecutive samples where the organism is moving with a similar speed and direction. However, since the time required for an organism to cross its home range is orders of magnitude larger than the time it moves with a consistent behavior (i.e., $\tau_p > \tau_v$), it is still possible to produce good estimates of range crossing time even if there is no information on the organism's speed. Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the organism's home-range sizes and home-range size, provided that the total sampling duration, $T$, is sufficiently long. Model selection procedures with criteria such as the Akaike Information Criterion [@akaike_new_1974] can help choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].

@@

### Quantifying home-range sizes

\noindent The concept of a home range has a long history in ecological research as an indicator of the space an organism requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for organisms which are not range-resident, such as migrating organisms [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic organisms [e.g., @morato_space_2016; @nandintsetseg_variability_2019]. Although here we quantify organism's home-range sizes using 95% home-range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of organisms that are not range-resident. In the case of organisms that are range resident for short periods of time, we can model their movement and short-term home-range sizes within distinct and discrete periods or using a sliding window approach (see the discussion section). Note, however, that although range-resident organisms will generally have smaller positional variance ($\vpos$) than migratory and nomadic organisms, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an organism's movement but not its behavior. Instead, the organism's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different behaviors. Organisms with a $\tau_p$ shorter than a year are likely range-resident, while organisms with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year). In contrast, when an organism's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it does not have an average position it returns to regularly.
-->

<!-- ### Modeling home-range sizes -->

@@

This section of the appendix illustrates the steps necessary to reproduce the tapir movement analysis and the related figure in the manuscript (fig. 5). The tapir data used here is from the work of @medici_movement_2022 and can be found at the GitHub repository located at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs). Estimating the effects of resource abundance and unpredictability on the tapir's home-range size requires us to first estimate the changes in the tapir's home-range sizes (section \@ref(movement)) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi)) before we can estimate the relationship between resource dynamics and space use (section \@ref(final-fig)).

@@

```{r, eval=TRUE, warning=FALSE, echo=FALSE}
# attach all necessary packages
library('terra')     # to import and save rasters
library('dplyr')     # for data wrangling
library('purrr')     # for functional programming
library('tidyr')     # for data wrangling
library('ggplot2')   # for fancy plots
library('cowplot')   # for fancy multi-panel plots
library('ctmm')      # for movement modeling
library('mgcv')      # for empirical Bayesian GAMs
library('lubridate') # for smoother date wrangling
library('sf')        # for spatial features
library('MODIStsp')  # for downloading NDVI rasters
library('dagitty')   # for directed acyclical graphs
library('ggdag')     # for directed acyclical graphs
library('gratia')    # for ggplot-based GAM figures
theme_set(theme_bw()) # change default theme
source('../functions/betals.r') # betals family written by Simon Wood
source('../analysis/figures/default-figure-styling.R') # for color palettes
source('../earthdata-login-info.R') # personal login info for EarthData
source('../functions/window_hr.R') # function to calculate HRs
```

@@

# axis labels (unicode characters removed because they knitting errors)
e_r <- bquote(paste(bold('Resource abundance')))
v_r <- bquote(paste(bold('Resource stochasticity')))
hr_lab <- bquote(bold(paste('7-day home-range size (', km^2, ')')))

@@

p_f <-
  expand_grid(mu = seq(from = floor(min(tapir$mu) * 100) / 100,
                       to = ceiling(max(tapir$mu) * 100) / 100,
                       length.out = 250),
              sigma2 = seq(from = 0, to = 0.0035, length.out = 250)) %>%
  mutate(hr_full_est = predict(m, newdata = ., type = 'response')) %>%
  ggplot() +
  geom_raster(aes(mu, sigma2, fill = hr_full_est)) +
  geom_contour(aes(mu, sigma2, z = hr_full_est), color = 'black') +
  geom_point(aes(mu, sigma2), tapir, alpha = 0.3, show.legend = FALSE) +
  scale_x_continuous(e_r, expand = c(0, 0)) +
  scale_y_continuous(v_r, expand = c(0, 0)) +
  scale_fill_gradient(bquote(atop(bold('7-day home-range'),
                                  paste(bold('size (km'), '\U00B2',
                                        bold(')')))),
                      low = 'grey90', high = pal[3], limits = c(0, NA)) +
  theme(legend.position = c(1, 1),
        legend.justification = c('right', 'top'),
        legend.box.background = element_rect(),
        legend.background = element_rect(),
        legend.key.width = unit(0.35, 'in')) +
  guides(fill = guide_colorbar(
    title.position = 'top',
    theme = theme(legend.title = element_text(hjust = 1)),
    direction = 'horizontal'))
